**Title:** Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs

### 2020-07-28 manuscript/ecosphere_resubmit/ecosphere_response.Rmd
making progress on edits

While we have full seasonal coverage for the majority of lake-years we decided to follow the reviewers suggestion and filter these out.  This filtering step has been added to our `data_clean.R` script and we remove any lake/year/parameter that does not have a minimum of one sample in May/June, July/August, and September/October for both the Watershed Watch data and the LAGOSNE data.  The changes to the analysis are minimal and do not alter any of our conclusions.

@@

- **Response:**

@@

- **Response:**

*1-4.* It is unclear how the trends in temperature, TP, chl-a, TN, and TN:TP compare to water quality condition and any standards that might be set in Rhode Island. Tables 1 and 2 give the range of values across lakes and there is clearly a large gradient in trophic status. Following from the comment made in point 8, it would be helpful to better understand the trends observed here in context of water quality condition. If chl-a is rising, is it still relatively low, or is it reaching “polluted/degraded” levels? Do you observe a shift in trophic status in individual lakes? Somehow, try to put the trends or lack of trends observed in the context of how water quality condition is changing over time. Line 312: It is still unclear to me how this analysis of site-specific anomalies is robust to variation in timing of inclusion of given sampling locations.

- **Response:**

@@

- **Response:**

@@

- **Response:**

@@

- **Response:**

@@

- **Response:**

@@

- **Response:**

@@

- **Response:**

@@

- **Response:**

@@

- **Response:**

@@

- **Response:**

@@

- **Response:**

@@

- **Response:**

@@

- **Response:**

>>>>

While we have full seasonal coverage for the majority of lake-years we decided to follow the reviewers suggestion and filter these out.  This filtering step has been added to our `data_clean.R` script at lines 163 and 259 and is in the `filter_months()` function in `functions.R`.  This removes any lake/year/parameter that does not have a minimum of one sample in May/June, July/August, and September/October for both the Watershed Watch data and the LAGOSNE data.  The changes to the analysis are minimal and do not alter any of our conclusions.  We added text in the second paragraph of the Methods section describing the selection criteria.

@@

- **Response:**  Similar to the approach in Oliver et al. we have added an additional step to ensure that lakes are only included if they have data in the first and second half of the 1993 - 2016 time frame and now only lakes that have at least one measurement in 1993-2004 and 2005-2016, have a minimum of 10 years of data, and have at least 1 measurement in may/june, july/aug, sept/oct are included in the analysis.  The filtering step has been added to the `data_clean.R` around line 233 and 316 and uses a new function, `filter_early_late()` which is in the `functions.R` file.  The impact on the analysis was minimal.  We added text in the second paragraph of the Methods section describing the selection criteria.

@@

- **Response:** As we were comparing the results from our analysis in Rhode Island to the LAGOSNE data and specifically to the results found in the Oliver et al 2017 paper, we constrained the analysis to a similar set of parameters.  The Oliver et al 2017 paper did not include secchi in their analysis and we decided to follow suit.

*1-4.* It is unclear how the trends in temperature, TP, chl-a, TN, and TN:TP compare to water quality condition and any standards that might be set in Rhode Island. Tables 1 and 2 give the range of values across lakes and there is clearly a large gradient in trophic status. Following from the comment made in point 8, it would be helpful to better understand the trends observed here in context of water quality condition. If chl-a is rising, is it still relatively low, or is it reaching “polluted/degraded” levels? Do you observe a shift in trophic status in individual lakes? Somehow, try to put the trends or lack of trends observed in the context of how water quality condition is changing over time.

- **Response:** NEEDS A RESPONSE!

*1-4a.* Line 312: It is still unclear to me how this analysis of site-specific anomalies is robust to variation in timing of inclusion of given sampling locations.

- **Response:** NEEDS A RESPONSE!

@@

- **Response:** We address the reviewers concern regarding the possible association between rising temperature and increasing chlorophyll in the individual pieces below.  In short, we concur and have done all we can to avoid the appearance of suggesting causation.

@@

- **Response:** NEEDS A RESPONSE!

@@

- **Response:** We like the reviewers concept of a more holistic way of interpreting trends.  We attempted to do this in our initial assessment of these trends, but we obviously were not successful in doing so.  We now lay out a more explicit approach and describe how we use sign of the slope, magnitude of the slope, relative frequency of high/low years in the beginning and and end of the time frame, and the p-value together to assess for trends.  This is described in the paper in a new section in the methods, *Assessing regressions for trends*.   One of the biggest changes is to suggest values for the slope that would indicate an ecologically meaningful trend.  We describe this in the new section, but in general we are using a trophic state change from oligotrophic to eutrophic over one century to indicate ecologically meaningful change.  For temperature we look at values associated with ecological change as well as policy values (e.g. the Paris Agreement).  While these exact numbers are not universal, we use them as a starting point for assesing the trends.  How we interpreted the trends relative to the additional specific issues raised by the reviewer are detailed below.

@@

- **Response:**  The results for TN in Rhode Island have changed somewhat since we added the additional filtering steps suggested by Reviewer 1 (*1-1* and *1-2*).  This weakened the positive trend somewhat as the slope is smaller and the p-value larger.  Additionally two years, 1995 and 1996, now have insufficient data to be included in the figures.  Using the more explicit definition of a trend we not conclude that TN showed a weak statistically positive trend but not an ecologically important trend over the time frame of a century as the slope is less than the amount needed to cause an trophic state change.  This is reflected in the manuscript.

@@

- **Response:** We applied the same within and across year filters to the LAGOSNE and these results have changed as well, largely becuase nutrient data across the full time frame was scarce in LAGOSNE in more recent years.  Given that and our method for interpreting the trends leads to the following:  LAGOSNE is showing weak statistical support for a negative change in nitrogen, more below average years recently.  The slope is less than the 3.5 we identified, albeit close.  Thus we conclude a weak decrease in TN at the scale of the Region.  This is indicated in the manuscript.

@@

- **Response:** Again, with the changes to the filtering the trend lines have slightly changed.  We concur with Reviewer 1's take that TP is showing a slight increasing trend with some statistical support for a trend, a value greater than the threshold we identified and greater frequency of above average years in recent years.  We have indicated as such in the text.

@@

- **Response:** This is a good suggestion, but we have chosen not to implement it, primarily because of Reveiwer 1's suggestion to have a more holistic approach to addressing the trends.  In developing our approach to assessing the trends we now have ecologically relevant thresholds we can compare to.  Both approaches address the same issue of putting the magnitude of the slopes in context.

@@

- **Response:**  We have softened our language in the discussion on original lines 236-237 considerably.  They now read, "While our analysis is not capable of detecting causation, this association between increasing chlorophyll *a* and warming waters is interesting and warrants further exploration to see if increasing chlorophyll *a* can be described by temperature alone."

@@

- **Response:** NEEDS A RESPONSE!

```{r, echo=FALSE}
x <- ww_data %>% 
  select(station_name, site_descr, town, year, param, station_year_mean) %>%
  spread(param, station_year_mean) %>%
  filter(complete.cases(.))
xyr <- group_by(x, station_name) %>% 
  summarize(yrly_mean_chla = mean(chla),
            yrly_mean_temp = mean(temp),
            yrly_mean_np = mean(np_ratio),
            yrl_mean_tn = mean(total_n),
            yrly_mean_tp = mean(total_p))
```

@@

- **Response:** We have added a new analysis to look at the same plots across trophic states.

@@

- **Response:** Fixed

@@

- **Response:** Fixed.
